prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-09 inst/MyArticle/paper.Rmd
merge from cp. Remove fold changes.

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analisis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level fold change and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage known from `r BiocStyle::Biocpkg("limma")`. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

@@

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. 
However, if there are no observation in one of the groups, then contrasts involving this group can not be estimated. To address this, for the group without observations, we use a point prior equal to the average of $3\%$ smallest group intensities means in the dataset and infer the variance using the pooled the variance of other groups. Proteins for which the fold change was imputed in such a way are explicitly annotated in the result table. 
Other means of modelling missing information in proteomics data using the Hurdle model is discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020]. 
We show using a benchmark dataset that this, although trivial approach, can outperform methods based on data imputation or dropout models.

@@

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. As a result, we obtain different fold changes by comparing the various dilutions, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected fold change is 1.2 for E. coli proteins and 1 for human proteins.
We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by MaxQuant are available in the _prolfquadata_ R package [@prolfquadata].

@@

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A) for the largest number of proteins (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated fold changes are unbiased (not shown) and show high diagnostic accuracy (\@ref(fig:FDRfdp) Panel A). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the linear model with variance moderation. Because of the robustness of the dropout model to missing observations, we intend to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the pAUC, and if the FDR is an unbiased estimate of the FDR. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice. First, estimate protein abundances from peptide abundances using a robust or nonparametric regression method. Second, use linear models as long as they suffice to model your experimental design because they show good performance in all categories. Third, use mixed effect models only if you have to, e.g., repeated measurements and if the sample sizes are large. If you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change. Last but not least, do not impute missing data but model it instead.

@@

\begin{table}[H]
\centering
\begin{tabular}[H]{ll}
\toprule
Abbreviations & Explaination\\
\midrule
AUC    &Area Under the Curve                               \\
CDF    &Cumulative Distribution Function                   \\
ESI-MS &Electro-Spray-Ionization Mass Spectrometry         \\
FC     &Fold Change                                        \\
$FDP$  &False Discovery Proportion                         \\
$FDR$  &False Discovery Rate                               \\
HPLC   &High-Pressure Liquid Chromatography                \\
LC-MS  &Liquid Chromatography followed by Mass Spectrometry\\
LFQ    &Label-Free Quantification                          \\
LOD    &Limit Of Detection                                 \\
MS     &mass spectrometry                                  \\
OO     &Object-Oriented                                    \\
UML     &       Unified Modeling Language                  \\
\bottomrule
\end{tabular}
\end{table}

@@

Here we compare the benchmark results for the `r BiocStyle::Biocpkg("proDA")` model and the linear model with variance moderation. We see that the fold change estimates obtained with the `r BiocStyle::Biocpkg("proDA")` package show better diagnostic accuracy.  However, the $t-statistics$ or $p-values$ show slightly worse diagnostic accuracy (Figure \@ref(fig:proDABenchmarks). Figure \@ref(fig:proDAAllBenchmarks), compares the fold change estimates obtained using both methods.

@@

(ref:proDAAllBenchmarks) Comparison of fold change estimates produced by proDA and produced by linear models implemented in prolfqua.

>>>>

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analisis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage known from `r BiocStyle::Biocpkg("limma")`. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

@@

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. 
However, if there are no observation in one of the groups, then contrasts involving this group can not be estimated. To address this, for the group without observations, we use a point prior equal to the average of $3\%$ smallest group intensities means in the dataset and infer the variance using the pooled the variance of other groups. Proteins for which the difference was imputed in such a way are explicitly annotated in the result table. 
Other means of modelling missing information in proteomics data using the Hurdle model is discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020]. 
We show using a benchmark dataset that this, although trivial approach, can outperform methods based on data imputation or dropout models.

@@

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. As a result, we obtain various effect sizes by comparing the various dilutions, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is 1.2 for E. coli proteins and 1 for human proteins.
We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by MaxQuant are available in the _prolfquadata_ R package [@prolfquadata].

@@

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A) for the largest number of proteins (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated differences are unbiased (not shown) and show high diagnostic accuracy (\@ref(fig:FDRfdp) Panel A). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the linear model with variance moderation. Because of the robustness of the dropout model to missing observations, we intend to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the pAUC, and if the FDR is an unbiased estimate of the FDR. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice. First, estimate protein abundances from peptide abundances using a robust or nonparametric regression method. Second, use linear models as long as they suffice to model your experimental design because they show good performance in all categories. Third, use mixed effect models only if you have to, e.g., repeated measurements and if the sample sizes are large. If you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference. Last but not least, do not impute missing data but model it instead.

@@

\begin{table}[H]
\centering
\begin{tabular}[H]{ll}
\toprule
Abbreviations & Explaination\\
\midrule
AUC    &Area Under the Curve                               \\
CDF    &Cumulative Distribution Function                   \\
ESI-MS &Electro-Spray-Ionization Mass Spectrometry         \\
$FDP$  &False Discovery Proportion                         \\
$FDR$  &False Discovery Rate                               \\
HPLC   &High-Pressure Liquid Chromatography                \\
LC-MS  &Liquid Chromatography followed by Mass Spectrometry\\
LFQ    &Label-Free Quantification                          \\
LOD    &Limit Of Detection                                 \\
MS     &mass spectrometry                                  \\
OO     &Object-Oriented                                    \\
UML     &       Unified Modeling Language                  \\
\bottomrule
\end{tabular}
\end{table}

@@

Here we compare the benchmark results for the `r BiocStyle::Biocpkg("proDA")` model and the linear model with variance moderation. We see that the difference estimates obtained with the `r BiocStyle::Biocpkg("proDA")` package show better diagnostic accuracy.  However, the $t-statistics$ or $p-values$ show slightly worse diagnostic accuracy (Figure \@ref(fig:proDABenchmarks). Figure \@ref(fig:proDAAllBenchmarks), compares the difference estimates obtained using both methods.

@@

(ref:proDAAllBenchmarks) Comparison of difference estimates produced by proDA and produced by linear models implemented in prolfqua.
